Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-01-21 papers/ArticleZ80/source/paper.tex
iniciando correções do linguisticas sugeridas por um professor de ingles.

\title{Experiences in Modelling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}
%
%
%
\author{Val\'{e}rio Medeiros Jr\inst{1}, David D\'{e}harbe\inst{2}}

@@

The rigorous development of an embedded software in a microcontroller requires 
a specification of its instruction set and the software requirements specification. 
The actual industrial practice is to report the instruction set informally using natural language
or with added semi-formal notation. This practice may leave space for errors and inconsistencies \cite{springerlink:Yuan2011}.
This paper presents specifically a B formal model of the Z80 instruction set. 
This model  can be  used on documentation, simulation, test and  a model-driven approach
to design and develop software from the functional specification level down to assembly.
%Using this model-driven approach and the instruction set model, a small case study to
%analyse a petroleum production test system in oil field was developed.
%  
% \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

In general, the manuals of microcontrollers can be more suited for \marginpar{DD: o título da sub-seção não condiz com seu conteúdo. Não cita nenhum problema na formalização. Não vejo muito bem o que ele traz de novo que já não foi dito anteriormente.}
developers in assembly language. Many times, each instruction of
microcontrollers is shown in its official manual using textual
description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. This
does not allow the interpretation by parser.  The manuals also have
semi-formal and informal descriptions, these descriptions may make
mistakes.  The official manual of the microcontroller Z80 has also
several problems that were identified over
time~\cite{UndocumentedZ80}.

@@

Usually, each type module just needs to specialize definitions that
are given in the hardware modeling library.  For example, the function
$\textit{bv\_to\_nat}$ from bit vector arithmetic is specialized to
$\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can defined as follows:
$$
\begin{array}{l}
\textit{byte\_uchar} \in \textit{BYTE} \fun \nat \land \\
\textit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$$

@@

This section presents the verification of the aforementioned modules
using the AtelierB provers.  By default, AtelierB~\cite{atelierB} does
not generate proof obligations of the type of functions defined in the
clause $PROPERTIES$ of a machine. For example, the definition $inc \in
\{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x + 1)$,
does not create a proof obligation to check the consistency of the
definition. However, this definition is inconsistent since $inc(1)=2$
and $ 2 \not\in \{0,1\}$.

To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause, that cause
corresponding proof obligations to be generated and verified. However,
we have not been able to build the corresponding proofs using the base
of rules included in the free distribution of AtelierB.

A solution would be to carry out that verification with ProB
\cite{proB}, as it is able to verify the type of functions, but its
support is generally only suitable for finite functions. Since these
functions use the concept of sequence (an infinite set), we have not
succeded in showing type correctness of the conversion functions with
ProB\marginpar{Estou perdido: não definiu tipos especiais BYTE e BV16
  para poder usar o ProB?}. Eventually, we formalized new proof rules
to check these properties in AtelierB and used them in the interactive
prover of AtelierB.

@@

\textbf{We needed to verify the type of these conversion
  functions}.\marginpar{Não precisa repetir
  isto.} % There are twenty rules associated this properties
The conversion functions need to satisfy two properties: to be
bijective and equal to the reverse of the dual conversion
function. The three simplest proof obligations were verified
automatically with the AtelierB: (\it uchar\_schar\rm \ $\in$
\textit{UCHAR} $\bij$ \textit{SCHAR}), (\it schar\_uchar\rm \ $\in$
\textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it byte\_bv16\rm \ $=$ \it
bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

\item The proof obligation (\it byte\_uchar\rm \ $\in$ \textit{BYTE} $\bij$ \textit{UCHAR}) and its expansion in rule format is:\\ 
$\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) \rm $\in$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )  $\bij$  \rm 0 $\upto$ \rm 2\rm 5\rm 5 \\

\item The proof obligation (\it uchar\_byte\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{BYTE}) and its expansion in rule format is:\\ 
$\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 0$\upto$255  $\mid$  \rm \rm  [(\it v0  $\mod$  \rm 2)$\div$1,(\it v0  $\mod$  \rm 4)$\div$\rm 2\rm,(\it v0  $\mod$  \rm 8)$\div$\rm 4\rm,(\it v0  $\mod$  \rm 16)$\div$\rm 8\rm,(\it v0  $\mod$  \rm 32)$\div$\rm 1\rm 6\rm,(\it v0  $\mod$  \rm 6\rm 4)$\div$\rm 3\rm 2\rm,(\it v0  $\mod$  \rm 1\rm 2\rm 8)$\div$\rm 6\rm 4\rm,(\it v0  $\mod$  \rm 2\rm 5\rm 6)$\div$\rm 1\rm 2\rm 8\rm \rm ]\rm ) \rm   $\in$  \rm 0 $\upto$ \rm 2\rm 5\rm 5   $\bij$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )\\

\item  The proof obligation (\it byte\_uchar\rm  =  \it uchar\_byte\rm$^{-1}$) and its expansion in rule format is:\\ 
  $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm )=($\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 0$\upto$255  $\mid$  \rm \rm  [(\it v0  $\mod$  \rm 2)$\div$1,(\it v0  $\mod$  \rm 4)$\div$\rm 2\rm,(\it v0  $\mod$ \rm 8) $\div$\rm 4\rm,(\it v0  $\mod$  \rm 16)$\div$\rm 8\rm,(\it v0  $\mod$  \rm 32)$\div$\rm 1\rm 6\rm,(\it v0  $\mod$ \rm 6\rm 4)$\div$\rm 3\rm 2\rm,(\it v0  $\mod$  \rm 1\rm 2\rm 8)$\div$\rm 6\rm 4\rm,(\it v0  $\mod$  \rm 2\rm 5\rm 6)$\div$\rm 1\rm 2\rm 8\rm \rm ]\rm )\rm)$^{-1}$
\end{enumerate}

Finally, four proof obligations are using rules to demonstrate that
the functions are finite\marginpar{Qual o objetivo de provar que as funções são finitas?}
: \it byte\_uchar $ \in \finset$\rm (\it
byte\_uchar\rm ); \it uchar\_byte $ \in \finset$\rm (\it
uchar\_byte\rm ); \it ushort\_bv16 $ \in \finset$\rm (\it
ushort\_bv16\rm ) and \it bv16\_ushort $ \in \finset$\rm (\it
bv16\_ushort\rm ).
\begin{enumerate}
\item  The proof obligation (\it byte\_uchar $\in$   $\finset$ \rm (\it byte\_uchar\rm )) and its 
expansion in rule format is\marginpar{Não entendi esta expansão.}:\\
  $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) $\in$   $\finset$ \rm ( $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm )\rm )  \\

\end{enumerate}

We have included twenty four rules related to binary arithmetic
as\marginpar{Qual a motivação para provar estas propriedades? Parece
  estar fora do escopo do trabalho apresentado no artigo.}  developed
in~\cite{Leibnizens}. These rules are used in the interactive theorem
prover with binary arithmetic to prove important
properties~\cite{James2010,Dasgupta2006}.

@@

	\bf SETS\\
	\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0 \rm , \it f0 \rm ,\it b0 \rm , \it c0 \rm ,\it d0 \rm , \it e0 \rm ,\it h0 \rm , \it l0 \rm ,\it a\_0 \rm , \it f\_0 \rm ,\it b\_0 \rm , \it c\_0 \rm ,\it d\_0 \rm , \it e\_0 \rm ,\it h\_0 \rm , \it l\_0 \rm \};\\
	\hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{ \it BC \rm , \it DE \rm , \it HL \rm , \it SP \rm , \it AF \rm \}

@@

The modeling of instruction set, interruptions and input/output ports become the animation interesting.
The animation in ProB version is not so fast and practical as a simulation, for example in \cite{Simulator_z80},
because ProB works manipulating logic and math concepts that can have infinite sets and non determinism, then ProB requires more processing, specially when it works analysing  complex expressions or very large expressions.
However, a user of ProB with this modeling can: execute manually a sequence of Z80's instructions that represent a program and view the visited states of ports, memory and registers, it is important to debug the effect action on the state space computed; create a tracing of the states; analyse constants, functions and expressions; and search possible violations.

@@

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The next works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small example
software verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process.
%Link com o parágrafo anterior 
These works created a important step to software verification up to assembly
language, they show the actuals difficulties and suggest improvements in the
B tools.

@@

Future works comprise the development of software with the B method from
functional specification to assembly level, using the Z80 model presented in this
work. The mechanic compilation from B algorithmic constructs to assembly platform
is also envisioned. Finally, another important activity is develop a formal model of a platform used in LLVM Compiler~\cite{DBLP:conf/cgo/LattnerA04}.

>>>>

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}
%
%
%
\author{Val\'{e}rio Medeiros Jr\inst{1}, David D\'{e}harbe\inst{2}}

@@

The rigorous development of an embedded software in a microcontroller requires a
specification of its instruction set and the software requirements specification.
The actual industrial practice is to document the instruction set informally
using natural language or with added semi-formal notation. This practice leaves
space for errors and inconsistencies \cite{springerlink:Yuan2011}. This paper
presents specifically a B formal model of the Z80 instruction set. This model 
can be  used to document, simulate, test and in a model-driven approach to
design and develop software from the functional specification level down to
assembly.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
%  \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

Many times, each instruction of microcontrollers is shown in its official manual
using textual description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. 
The manuals also have semi-formal and informal descriptions, these descriptions
may leave errors, inconsistency and ambiguity.  
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users \marginpar{VM: Adicionei um exemplo real e uma referência  que cita vários problemas identificados no Manual do Z80.} 
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description, that  sometimes can be placed in different pages. 
Furthermore, the official manual of the microcontroller Z80 has also several
problems that were identified over time~\cite{UndocumentedZ80}.

@@

Usually, each type module just needs to specialize definitions that
are given in the hardware modeling library.  For example, the function
$\textit{bv\_to\_nat}$ from bit vector arithmetic is specialized to
$\textit{byte\_uchar}$. As the set $\textit{BYTE}$ is a subset of the
$\textit{BIT\_VECTOR}$, this function can be defined as follows:
$$
\begin{array}{l}
\textit{byte\_uchar} \in \textit{BYTE} \fun \nat \land \\
\textit{byte\_uchar} = \lambda (v) . ( v \in BYTE | bv\_to\_nat(v) )
\end{array}
$$

@@

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does
not generate proof obligations of the type of functions defined in the
clause $PROPERTIES$ of a machine. For example, the definition $inc \in
\{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x + 1)$,
does not create a proof obligation to check the consistency of the
definition. However, this definition is inconsistent since $inc(1)=2$
and $ 2 \not\in \{0,1\}$.

To avoid such inconsistencies, we state separately the required properties in the
\textit{ASSERTIONS\/} clause, that cause corresponding proof obligations to be
generated and verified. However, we have not been able to build the corresponding
proofs using the base of rules included in the free distribution of AtelierB. So
we formalized new proof rules to check these properties in AtelierB and used them
in the interactive prover of AtelierB.

@@

%\textbf{We needed to verify the type of these conversion
%  functions}.\marginpar{Não precisa repetir
%  isto.} % There are twenty rules associated this properties
The conversion functions need to satisfy two properties: to be
bijective and equal to the reverse of the dual conversion
function. The three simplest proof obligations were verified
automatically with AtelierB: (\it uchar\_schar\rm \ $\in$
\textit{UCHAR} $\bij$ \textit{SCHAR}), (\it schar\_uchar\rm \ $\in$
\textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it byte\_bv16\rm \ $=$ \it
bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

\item For the proof obligation (\it byte\_uchar\rm \ $\in$ \textit{BYTE} $\bij$
\textit{UCHAR}), the proof rule is:\\
$\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) \rm $\in$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )  $\bij$  \rm 0 $\upto$ \rm 2\rm 5\rm 5 \\

\item For the proof obligation (\it uchar\_byte\rm \ $\in$ \textit{UCHAR} $\bij$
\textit{BYTE}), the proof rule is:\\ $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 0$\upto$255  $\mid$  \rm \rm  [(\it v0  $\mod$  \rm 2)$\div$1,(\it v0  $\mod$  \rm 4)$\div$\rm 2\rm,(\it v0  $\mod$  \rm 8)$\div$\rm 4\rm,(\it v0  $\mod$  \rm 16)$\div$\rm 8\rm,(\it v0  $\mod$  \rm 32)$\div$\rm 1\rm 6\rm,(\it v0  $\mod$  \rm 6\rm 4)$\div$\rm 3\rm 2\rm,(\it v0  $\mod$  \rm 1\rm 2\rm 8)$\div$\rm 6\rm 4\rm,(\it v0  $\mod$  \rm 2\rm 5\rm 6)$\div$\rm 1\rm 2\rm 8\rm \rm ]\rm ) \rm   $\in$  \rm 0 $\upto$ \rm 2\rm 5\rm 5   $\bij$ \rm (\rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}\rm )\\

\item For the proof obligation (\it byte\_uchar\rm  =  \it
uchar\_byte\rm$^{-1}$), the proof rule is:\\ $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm )=($\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 0$\upto$255  $\mid$  \rm \rm  [(\it v0  $\mod$  \rm 2)$\div$1,(\it v0  $\mod$  \rm 4)$\div$\rm 2\rm,(\it v0  $\mod$ \rm 8) $\div$\rm 4\rm,(\it v0  $\mod$  \rm 16)$\div$\rm 8\rm,(\it v0  $\mod$  \rm 32)$\div$\rm 1\rm 6\rm,(\it v0  $\mod$ \rm 6\rm 4)$\div$\rm 3\rm 2\rm,(\it v0  $\mod$  \rm 1\rm 2\rm 8)$\div$\rm 6\rm 4\rm,(\it v0  $\mod$  \rm 2\rm 5\rm 6)$\div$\rm 1\rm 2\rm 8\rm \rm ]\rm )\rm)$^{-1}$
\end{enumerate}

Eight proof obligations require that conversion functions are finite. %: \it
% byte\_uchar $\in$ $\finset$\rm (\it byte\_uchar\rm ); \it uchar\_byte  $\in$
% $\finset$\rm (\it uchar\_byte\rm ); \it ushort\_bv16 $ \in \finset$\rm (\it
% ushort\_bv16\rm ); \it bv16\_ushort $ \in \finset$\rm (\it bv16\_ushort\rm ); \it
% byte\_schar $ \in \finset$\rm (\it byte\_schar\rm ); \it schar\_byte $ \in
% \finset$\rm (\it schar\_byte\rm ); \it sshort\_bv16 $ \in \finset$\rm (\it
% sshort\_bv16\rm ); and \it bv16\_sshort $ \in \finset$\rm (\it bv16\_sshort\rm).
 These proof obligations were verified using the expression evaluator of ProB.
Finally, we have included rules related to binary arithmetic as developed
in~\cite{Leibnizens}. These rules are used in the interactive theorem prover with
binary arithmetic to prove important properties~\cite{James2010,Dasgupta2006}.

% Finally, four proof obligations  are using rules to demonstrate that
% the functions are finite\marginpar{Qual o objetivo de provar que as funções são finitas?}
% : \it byte\_uchar $ \in \finset$\rm (\it
% byte\_uchar\rm ); \it uchar\_byte $ \in \finset$\rm (\it
% uchar\_byte\rm ); \it ushort\_bv16 $ \in \finset$\rm (\it
% ushort\_bv16\rm ) and \it bv16\_ushort $ \in \finset$\rm (\it
% bv16\_ushort\rm ).
% \begin{enumerate}
% \item  For the proof obligation (\it byte\_uchar $\in$   $\finset$ \rm (\it byte\_uchar\rm )) , the proof rule is:\\
%   $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm ) $\in$   $\finset$ \rm ( $\lambda$ \it v0\rm .\rm (\it v0 $\in$  \rm 1 $\upto$ \rm 8  $\fun$  \rm \{\rm 0\rm ,\rm 1\rm \}  $\mid$  \rm 1\rm 2\rm 8 $\times$ \it v0\rm (\rm 8\rm )\rm +\rm 6\rm 4 $\times$ \it v0\rm (\rm 7\rm )\rm +\rm 3\rm 2 $\times$ \it v0\rm (\rm 6\rm )\rm +\rm 1\rm 6 $\times$ \it v0\rm (\rm 5\rm )\rm +\rm 8 $\times$ \it v0\rm (\rm 4\rm )\rm +\rm 4 $\times$ \it v0\rm (\rm 3\rm )\rm +\rm 2 $\times$ \it v0\rm (\rm 2\rm )\rm +\rm 1 $\times$ \it v0\rm (\rm 1\rm )\rm )\rm )  \\
% 
% 
% \end{enumerate}

@@

	\bf SETS\\
\hspace*{0.10in}\it id\_reg\_8 \rm = \rm \{ \it a0\rm, \it f0\rm, \it b0\rm, \it
c0\rm, \it d0\rm, \it e0\rm, \it h0\rm, \it l0\rm, \it a\_0\rm,  \it f\_0\rm,
\it b\_0\rm, \it c\_0\rm, \it d\_0\rm, \it e\_0\rm, \it h\_0\rm, \it l\_0 \rm
\};\\ \hspace*{0.10in}\it id\_reg\_16 \rm = \rm \{  \it AF \rm, \it BC\rm, \it
DE\rm, \it HL\rm, \it SP\rm\}

@@

The modeling of instruction set, interruptions and input/output ports become the
animation interesting. The animation in ProB is not so fast and practical as a
simulation, for example in \cite{Simulator_z80}, because ProB works manipulating
logic and math concepts that can have infinite sets and non determinism, then
ProB requires more processing, specially when it works analysing  complex
expressions or very large expressions. However, a user of ProB with this modeling
can: execute manually a sequence of Z80's instructions that represent a program
and view the visited states of ports, memory and registers. It is important to
debug the effect action on the state space computed; create a tracing of the
states; analyse constants, functions and expressions; and search possible
violations.

@@

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small 
software that was verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process.
%Link com o parágrafo anterior 
These works created an important step to software verification up to assembly
language, they show the current difficulties and suggest improvements in the
B tools.

@@

Future works comprise the development of software with the B method from
functional specification to assembly level, using the Z80 model presented in this
work. The mechanic compilation from B algorithmic constructs to assembly platform
is also envisioned. Finally, another important activity is to develop a formal
model of a platform used in LLVM Compiler~\cite{DBLP:conf/cgo/LattnerA04}.
